Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-03 _tex/index.tex
Built site for gh-pages

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
be suitable for applications beyond the psycho-physical contexts for
which it was created. Thurstone himself cautioned that its use ``should
not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
could prove problematic in the presence of complex traits or
heterogeneous stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve these traits and stimuli, two main
assumptions of Case V may not consistently hold in theory or practice:
the zero correlation and equal dispersion between stimuli.

@@

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, consequently, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
disproportionate emphasis on grammatical accuracy, ultimately favoring
texts with fewer errors but weaker arguments. While direct evidence for
this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures in CJ data,
the statistical handling of this extra source of dependency has been
inadequate. For example, when CJ data includes multiple samples of
stimuli from the same individuals, researchers often rely on (average)
estimated BTL scores to conduct subsequent analyses and tests at the
individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
importance of addressing factors that create dependencies between
stimuli, as failing to do so can lead to inaccurate conclusions about a
trait's reliability and, by extension, its validity. For instance,
neglecting additional traits relevant to the stimuli, such as judges'
biases, often leads to a dimensional mismatch in the model. This
mismatch can potentially inflate the reliability of the trait
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias into the
trait's estimates \citep{Ackerman_1989}. Similarly, failing to account
for hierarchical (grouping) structures reduces the precision of model
parameter estimates, further amplifying the overestimation of
reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
undermine the trait's reliability, and because ``validity cannot exist
without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
compromise the validity of results derived from the the data modeling.

Fortunately, the same literature provides solutions for addressing the
factors that create dependencies between stimuli. \citet{Andrich_1978}
and \citet{Wainer_et_al_1978} propose integrating judges' biases into
the CJ model. This integration would produce a model similar to a
multidimensional or two-parameter logistic IRT model
\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
assumptions a researcher is willing to make regarding the judges'
biases. The literature also suggests incorporating relevant hierarchical
structures into the model to account for dependencies in the stimuli,
resulting in a model resembling a Multilevel Structural Equation Model
(MSEM) \citep[chap.~26]{Hoyle_et_al_2023}.

>>>>

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
be suitable for applications beyond the psycho-physical contexts for
which it was created. Thurstone himself cautioned that its use ``should
not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
could prove problematic in the presence of complex traits or
heterogeneous stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve these traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the zero correlation and equal
dispersion between stimuli.

@@

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, ultimately, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
disproportionate emphasis on grammatical accuracy, ultimately favoring
texts with fewer errors but weaker arguments. While direct evidence for
this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between the
stimuli has been inadequate. For example, when CJ data includes multiple
samples of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
importance of addressing factors that create dependencies between
stimuli, as failing to do so can lead to inaccurate conclusions about a
trait's reliability and, by extension, its validity. For instance,
neglecting additional traits relevant to the stimuli, such as judges'
biases, often leads to a dimensional mismatch in the model. This
mismatch can potentially inflate the reliability of the trait
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias into the
trait's estimates \citep{Ackerman_1989}. Similarly, failing to account
for hierarchical (grouping) structures reduces the precision of model
parameter estimates, further amplifying the overestimation of
reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
undermine the trait's reliability, and because ``validity cannot exist
without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
compromise the validity of results derived from the data modeling.

Fortunately, the same literature offers solutions for addressing the
factors that create these dependencies between stimuli.
\citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend integrating
judges' biases into the BTL model. Furthermore, the literature advocates
for incorporating relevant hierarchical structures into the data model
to account for these dependencies. Taken together, these additions can
result in a model that combines a Multilevel Structural Equation Model
(MSEM) \citep[chap.~26]{Hoyle_et_al_2023} with a multidimensional or
two-parameter logistic IRT model \citep[chap.~15]{Hoyle_et_al_2023},
depending on the theoretical assumptions a researcher makes about the
judges' biases.
